When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-11-03 paper/paper.tex
VF3 a bit differently

The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem. Note also that VF3 will sometimes crash on especially sparse graphs: we have
treated these instances as timeouts.

@@

For larger patterns of order 20 and 30, we have a large unsatisfiable region in the middle. Despite
not being near either phase transition, instances in the centre remain computationally challenging.
We also plot patterns of orders 14, 15 and 16, to show the change between the two behaviours.

We might expect these complexity plots to be symmetric along the diagonal, since for the induced
problem, if we replace both inputs with their complements, the solutions remain the same.  For the
Glasgow solver, this is almost the case. This should be expected, because this complement property is
precisely how the Glasgow solver handles the induced variant (although the heuristics may differ
between the two, which we discuss below). For LAD, some of the very dense patterns are slightly
harder than their diagonal opposites (LAD reasons about degrees, but not about complement-degrees);
it is interesting to note that for larger target graphs, VF2 finds \emph{all} dense pattern graphs
difficult. Finally, VF3's performance generally improves upon VF2's on instances that VF2 finds easy
(except in the cases where VF3 crashes), but worsens on instances that VF2 finds hard.

@@

\citet{o:CarlettiFSV17} correctly observe that ``a performance improvement on [hard instances] has a
far greater practical impact than it would have on easier graphs, since the saved time can be of
several hours or even days'', but mistakenly believe that being larger and slightly denser is what
makes an instance harder. To evaluate VF3, they introduce a new dataset consisting entirely of
satisfiable instances created by taking a connected subgraph of a randomly generated graph and
permuting the nodes. They conclude that VF3 is the best algorithm based upon its performance solving
the enumeration problem on these new instances.  We urge a more cautious interpretation of these
results: as we saw in \cref{section:induced}, being genuinely hard (rather than simply larger) is a
subtle property, and more care must be taken to avoid accidentally measuring which algorithm scales
best to large, easy instances due to it using the simplest inference routines. Indeed, we saw in
\cref{figure:induced} that VF3 performs \emph{worse} than VF2 on genuinely harder instances, and an
induced analogue of \cref{figure:labels} confirms that VF3 still struggles with many obviously
unsatisfiable instances.

>>>>

The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem.

@@

For larger patterns of order 20 and 30, we have a large unsatisfiable region in the middle. Despite
not being near either phase transition, instances in the centre remain computationally challenging
for every solver.  We also plot patterns of orders 14, 15 and 16, to show the change between the two
behaviours.

We might expect these complexity plots to be symmetric along the diagonal, since for the induced
problem, if we replace both inputs with their complements, the solutions remain the same.  For the
Glasgow solver, this is almost the case. This should be expected, because this complement property is
precisely how the Glasgow solver handles the induced variant (although the heuristics may differ
between the two, which we discuss below). For LAD, some of the very dense patterns are slightly
harder than their diagonal opposites (LAD reasons about degrees, but not about complement-degrees).

It is interesting to note that for larger target graphs, VF2 finds \emph{all} dense pattern graphs
difficult. Meanwhile, VF3 behaves better than VF2 in some regions, but worse than VF2 in others, and
has a larger hard region than either domain-based algorithm.\footnote{The version of
VF3 supplied by its inventors will crash if given an empty pattern graph. We have chosen to treat
these results as successes taking zero search, even though empty pattern graphs are not trivial for
the induced problem. If the reader prefers to interpret a crash as a timeout, the VF3 plots would
contain an additional area of black points down the left-hand side.} VF3 also occasionally finds
some instances with sparse target graphs (along the bottom of the plot) extremely difficult, whilst
other solvers do not.

@@

\citet{o:CarlettiFSV17} correctly observe that ``a performance improvement on [hard instances] has a
far greater practical impact than it would have on easier graphs, since the saved time can be of
several hours or even days'', but mistakenly believe that being larger and slightly denser is what
makes an instance harder. To evaluate VF3, they introduce a new dataset consisting entirely of
satisfiable instances created by taking a connected subgraph of a randomly generated graph and
permuting the nodes. They conclude that VF3 is the best algorithm based upon its performance solving
the enumeration problem on these new instances.  We urge a more cautious interpretation of these
results: as we saw in \cref{section:induced}, being genuinely hard (rather than simply larger) is a
subtle property, and it is not safe to say that an instance is hard simply because several
minor variations of a particular kind of algorithm find it hard. More care must thus be taken to
avoid accidentally measuring which algorithm scales best to large, easy instances due to it using
the simplest inference routines. Indeed, we saw in \cref{figure:induced} that VF3 performs
\emph{worse} than VF2 on genuinely harder instances, and additionally behaves erratically in parts
of the parameter space that every other solver finds easy; finally, an induced analogue of
\cref{figure:labels} confirms that VF3 also still struggles with many obviously unsatisfiable
instances.
